Prediction of Periprosthetic Joint Infection after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden and ...

### 2020-05-13 reports/article.Rmd
manuscript updates due to NH:s corrections. Fix #25.

# Extract predictors for specified period
# Needs additional formatting
predictors <- function(period) {
  tbl_coefs %>% 
  filter(
    time == period,
    term != "(Intercept)"
  ) %>% 
  transmute(
    term = tolower(term),
    term = replace(
      term, 
      startsWith(term, "diagnosis"), 
      "the underlying diagnosis for THA"
    ),
    term = replace(term, startsWith(term, "bmi"), "body mass index"),
    term = gsub("male ", "", term)
  ) %>% 
  pluck(1) %>% 
  unique() %>%
  glue::glue_collapse(", ", last = " and ")
}

@@

**RESULTS: ** The crude incidence rate of PJI after THA was `r crude_rate_90d * 100` % within 90, days and `r crude_rate_2y * 100` % within two years in the Swedish derivation cohort. A prediction model for PJI within 90 days combined `r predictors("90 days")` and had `r get_auc("90 days", "Reduced model", "2")` internally  and **(AUC=XXX, 95% CI: XXX to XXX)** externally. A similar model for PJI within two years had similar performance internally `r get_auc("2 years", "Reduced model")` and externally **XXX (95% CI: XXX to XXX)**. These models were superior to traditional models based on the American Society for Anesthesiologists (ASA) classification `r get_auc("90 days", "ASA class")`, Charlson `r get_auc("90 days", "Charlson")`, Elixhauser `r get_auc("90 days", "Elixhauser")`, and the Rx Risk V `r get_auc("90 days", "Rx Risk V")` co-morbidity indices.

@@

Periprosthetic joint infection (PJI) is the most devastating of all early complications after total hip arthroplasty (THA), mainly due to its severity in terms of increased mortality, the required number of re-operations, long-term antibiotic treatment, and often persistently impaired quality of life.[@Zimmerli2006; @Kapadia2016; @Shohat2019] Many risk factors for the development of PJI are identified, with anemia, diabetes, and obesity being important, to some extent modifiable risk factors, whereas advanced age and male gender are examples of non-modifiable risk factors.[@Ridgeway2005; @Lai2007; @Malinzak2009; @Ong2009; @Pedersen2010; @Bozic2012; @Jamsen2012; @Rasouli2014a; @Maoz2015]
Individualized PJI risk stratification would be a highly desirable tool in the pre-operative assessment of patients scheduled for THA, but established measures of co-morbidity such as the American Society for Anesthesiologists (ASA) classification or considerably more complex co-morbidity indices cannot be used in this context. Although high ASA class is a risk factor associated with development of PJI, the classification is imprecise, and it was never developed for the purpose of predicting this specific complication.[@Haynes1995; @Berbari2012] 
The Charlson[@Charlson1987] and Elixhauser[@Elixhauser1998] co-morbidity indices are based on `r nrow(coder::charlson)` or `r nrow(coder::elixhauser)` conditions coded by the International Classification of Diseases (ICD), respectively, but they are cumbersome to use in a clinical setting, and they were also not designed for prediction of specific adverse events after THA. The same is true for the Rx Risk V-classification with `r nrow(coder::rxriskv)` conditions codified by the Anatomical Therapeutic Chemical (ATC) Classification.[@Sloan2003; @Pratt2018]
Several attempts have been made at developing individualized risk prediction models for clinical usage, with the Universal American College of Surgeons National Surgical Quality Improvement Program (ACS NSQIP) Surgical Risk Calculator being one of the first.[@Edelstein2015; @Wingert2016a] Several other risk calculators for the risk of PJI after THA have since been proposed based on US or Australian cohorts, but none has gained general acceptance.[@Wingert2016a; @Bozic2013a; @Inacio2015a] This is mainly due to often limited, sometimes single-center based samples, low predictive and discriminative ability of the prediction models, or the absence of external model validation.
Thus, there is an urgent need for improved prediction models of PJI. We therefore developed a prediction model of PJI within 90 days or two years after THA that is based upon easily accessible data that are available in the setting of clinical decision-making.

@@

The Charlson, Elixhauser and Rx Risk V co-morbidity indices are based on the presence of ICD-10 and ATC codes grouped into main categories (`r nrow(coder::charlson)`, `r nrow(coder::elixhauser)` and `r nrow(coder::rxriskv)` respectively).[@Quan2005; @Pratt2018] Diagnostic categories from all indices were combined to define `r nrow(tab_categorization)` categories of co-morbidity that were chosen in order to be easily identifiable in a clinical setting (Tab. \@ref(tab:tabcategorization)). Some of the combined co-morbidities were observed for less than 10 patients either with or without PJI. These diagnoses were not considered as potential predictors to reduce the risk of overfitting to spurious events. PJI was defined by the occurrence of relevant ICD-10 or procedural NOMESCO-code recorded in NPR within 90 days or 2 years (Tab. \@ref(tab:infcodes)), or if the patient was reported to SHAR, within the same time frame, as reoperated due to infection.

@@

For each outcome (PJI within 90 days or 2 years), 1,000 bootstrap samples were drawn from the observed data set.[@Austin2004] Ten-fold cross validation was performed for every bootstrap sample with a range of potential penalty values ($\lambda$:s) in a logistic LASSO model, keeping only $\lambda$:s that minimized the mean cross-validated deviances in each sample and using these $\lambda$:s to estimate model coefficients for each potential predictor. Those coefficients were ranked based on their absolute values as a measure of  variable importance. Piece-wise linear regression was applied to find a break-point for which influential predictors had their estimated values above. Such predictors scored a point and the whole process was repeated 100 times. Potential predictors that were selected at least once were used in main effects multivariable logistic regression models without penalty and without pre-normalization of numeric variables. Even more parsimonious models with potential predictors chosen at least 33 out of 100 times were evaluated for comparison. Univariable prediction models based on either ASA class, the Charlson, Elixhauser, or the Rx Risk V co-morbidity indices were fitted for comparison. Odds ratios for the final models were estimated with 95 % confidence intervals.

@@

Each developed model was used to predict the probability of PJI within 90 days or two years for patients derived from the SHAR (internal validation). Receiver operating characteristic (ROC) curves and their corresponding areas under those curve (AUC) were used to describe discrimination with 95 % confidence intervals that were based on percentiles from 2,000 non-parametric bootstrap samples. The bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples was used to adjust for optimism. Predicted probabilities and observed proportions were plotted with 95 % confidence intervals to graphically assess model calibration.[@Nattino2016] The derived models were externally validated on a **XXX** cohort, and re-calibration of model intercepts was performed to account for different PJI incidences in Sweden compared to **XXX**. Calibration for this re-calibrated model was illustrated in the same calibration belt plot as after internal calibration.

@@

We built an online web calculator available at https XXX to be used in clinical practice. R software (version 3.6.1; R Foundation for Statistical Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr, pROC, decoder, coder and shiny were used.

@@

```{r}

males <- 
  sprintf(
    "%.0f %%, n = %s", 
    mean(df$P_Sex == "Male") * 100,
    format(sum(df$P_Sex == "Male"), big.mark = ",")
)

age <- sprintf("%.0f (SD = %.0f)", mean(df$P_Age), sd(df$P_Age))

@@

We included `r nmbr(N_shar)` patients (`r males` males) with mean age `r age` years in the derivation cohort from the SHAR (Fig. \@ref(fig:flowchart), left panel). 
`r crude_rate_90d * 100` % (n = `r sum(df$outcome_infection_90d)`) of the patients had PJI within 90 days, and `r crude_rate_2y * 100` % (n = `r sum(df$outcome_infection_2y)`) within 2 years (Tab. \@ref(tab:tab1)). The proportion of patients with ASA class III was
`r comorb(df$P_ASA == "III")`; `r comorb(df$CCI_index_quan_original > 0)` had at least one comorbidity according to the Charlson co-morbidity index, `r comorb(df$ECI_index_sum_all > 0)` according to Elixhauser, and `r comorb(df$Rx_index_pratt > 0)` according to Rx Risk V. In addition, **XXX** patients from **XXX **were included as an external validation cohort (Fig. \@ref(fig:flowchart), right panel).

@@

Patietns with `r excl_factors` were uncommon, wherefore those co-morbidities were excluded as potentisl predictors prioir to further modeling. ROC-curves (Fig. \@ref(fig:rocs)) and AUC-values (Fig. \@ref(fig:aucci)) for the main and reduced models were no different. We therefore focus on the reduced models due to simplicity. The correction for optimism only affected the third decimals in the presented results, and will therefore not be further discussed. Estimated probability density curves for patients with and without PJI were partially overlapping, although patients with PJI had on average higher predicted probabilites for such event (\@ref(fig:separation)). The reduced model for the prediction of PJI within 90 days in the Swedish cohort included `r predictors("90 days")` (Tab. \@ref(tab:coefs)). Predictors associated with a statistically significant risk increase were **AVN or secondary osteoarthritis as the underlying reason for THA surgery, the presence of CNS disease, obesity, and male gender**. The reduced model for the predicition of PJI within 90 days had `r get_auc("90 days", "Reduced model", "2")`, whereas ASA class `r get_auc("90 days", "ASA class")`, the Charlson co-morbidity index `r get_auc("90 days", "Charlson")`, the Elixhauser co-morbidity index `r get_auc("90 days", "Elixhauser")`, and Rx Risk V `r get_auc("90 days", "Rx Risk V")` had less discriminative power (Fig. \@ref(fig:rocs) and \@ref(fig:aucci)). Calibration was good for the reduced model (Fig. \@ref(fig:calibration), left panel).
The reduced model for PJI within two years included the same predictors, with the addition of the **presence of rheumatic disease** (Tab. \@ref(tab:coefs)). Predictors conferring a statistically significant risk increase were those **mentioned as predictive for PJI within 90 days, and additionally included the presence of psoriatic, rheumatic, or CNS disease, and cancer**. The reduced model for the prediction of PJI within two years had `r get_auc("2 years", "Reduced model", "2")`, whereas ASA class `r get_auc("2 years", "ASA class")`, the Charlson co-morbidity index `r get_auc("2 years", "Charlson")`, the Elixhauser co-morbidity index `r get_auc("90 days", "Elixhauser")`, and Rx Risk V `r get_auc("90 days", "Rx Risk V")` again had less discriminative power (Fig. \@ref(fig:rocs) and \@ref(fig:aucci)). Calibration was good for the reduced model (Fig. \@ref(fig:calibration), right panel).

@@

We found that a multivariable main effects logistic regression model based on some easily identifiable clinical conditions, gender, and BMI was considerably better at predicting PJI within 90 days after THA than models based on either ASA class or the Charlson, Elixhauser or Rx Risk V co-morbidity indices. A similarly developed model to predict PJI within two years included the same predictors as the model developed to predict PJI within 90 days, but added the presence of **rheumatic disease** to the list of predictor variables. This model also performed better than the established co-morbidity measures in terms of AUC and calibration.

@@

Since the prediction models both for PJI within 90 days and within two years were based on easily accessible information they can be translated into a simpole formula $\hat p = (e^{\hat \alpha + \mathbf{\hat \beta X}})/(1 + e^{\hat \alpha + \mathbf{\hat \beta X}})$ where $\hat p$ is the estimated probability, $\hat \alpha$ and $\mathbf{\beta} = \hat\beta_1, \hat\beta_2, \ldots, \hat\beta_k$ the estimated intercept and coefficients (Tab. \@ref(tab:coefs)) and $\mathbf{X} = X_1, X_2, \ldots, X_k$ a vector of corresponding patient characteristics. For example, `r examples$desc[[1]]` for PJI within 90 days; `r examples$desc[[2]]`.

@@

Patients with primary osteoarthritis seem less prone to developing PJI than patients receiving THA for other reasons, and in agreement with previous observational data we found that both AVN and secondary osteoarthritis conferred an about two-fold increased risk of PJI in our prediction models.[@Pedersen2010a; @Dale2012] Rheumatoid arthritis as the diagnosis underlying THA surgery is pointed out as a risk factor for PJI,[@Bozic2012; @Inacio2015a] but this predictor variable only reached borderline statistical significance in our model of 90-day PJI. However, the presence of psoriatic disease, and, for the outcome PJI within two years, the presence of rheumatic disease, were both statistically significant predictors of outcome in our models. This seeming discrepancy may be explained by the fact that other physicians may have identified patients as suffering from any of these conditions, thereby assigning the indicative ICD codes, whereas the operating surgeon may have classified the underlying reason for THA as something other than inflammatory joint disease. 
Obesity is associated with an increase in the risk of reoperations or adverse events after THA[@Sayed-Noor2019] and with the risk of developing surgical site infections or PJI after total hip or knee arthroplasty.[@Malinzak2009; @Jamsen2012; @Maoz2015; @Everhart2013] Obesity is identified as a risk factor for PJI in a risk calculator derived on two independent US institutional cohorts, which agrees with our identification of obesity as a predictor variable in both main models.[@Tan2018] In accordance with observational studies[@Jamsen2012] we found the presence of CNS diseases to be a predictor of PJI, and our rather broad category encompasses both cerebrovascular disease, dementia, hemiplegia, and Parkinson's disease which were all associated with a risk increase in the cited study. Cancer seems to confer an increased risk of PJI in several observational studies,[@Bozic2012; @Everhart2013; @Poultsides2013] and it is one of the defined co-morbidities in the ACS NSQIP Surgical Risk Calculator.[@Wingert2016a] Male gender is associated with an increased risk of developing PJI after arthroplasty surgery,[@Dale2012] and gender is included as a predictor variable in several previously developed risk calculators.[@Wingert2016a; @Tan2018] Taken together, the above cited findings support that the predictor variables in our models seem to be relevant in the context of predicting PJI.
High ASA class is frequently associated with an increased risk of PJI,[@Berbari2012; @Pulido2008; @Namba2012] as are co-morbidity in a broader sense, and, more specifically, the presence of cardiovascular co-morbidity.[@Ong2009; @Higuera2011] In contrast, we found that ASA class alone was an insufficient predictor of PJI, and none of the cardiovascular disease categories remained as a predictor variable in our final main model. Similarly, although mentioned as a risk factor for surgical site infection after joint arthroplasty,[@Rasouli2014a] our prediction models based on the Charlson co-morbidity index alone resulted in poor predictive power.
THA fixation using cement without antibiotics confers an increased risk of PJI,[@Dale2012] but such cement brands were not in clinical use in Sweden during the studied period. We found no indication that the type of fixation was associated with the risk of PJI, and this variable was also not selected as a potential predictor variable in any of the iterations during model development.

@@

Our study is hampered by the limitations inherent to observational research, such as misclassification, missing information, and XXX.
The presence of diabetes was not among the predictor variables selected in our models, although it was almost double as frequent among patients with PJI than among those without (data not shown). This is in disagreement with numerous observational studies describing an increased risk of developing PJI after both THA or total knee arthroplasty,[@Pedersen2010; @Maoz2015; @Everhart2013] and diabetes is also found among the risk factors in previously developed risk calculators.[@Wingert2016a; @Tan2018; @Inacio2015] On the other hand, diabetes is not consistently found to be associated with the risk of surgical site infection or PJI after THA surgery.[@Namba2012] Peri-operative hyperglycemia is a predictor of PJI after total joint replacement of the hip or knee, and information on laboratory parameters such as morning blood glucose or HBA1c could have improved the predictive power of our models, but we had no access to such information.[@Mraovic2011; @Iorio2012] In analogy, low preoperative hemoglobin is associated with the risk of developing surgical site infection,[@Rasouli2014a] but although we had access to the ICD codes defining the presence of anemia, we had no information on actual laboratory findings. 
Use of TNF-a blockers may be associated with the risk of developing PJI,[@Gilson2010; @Momohara2011] but we had no information on medication. Similarly, intra-articular steroid application may enhance the risk of developing PJI,[@Cancienne2015] but such manipulations would not be identifiable within a Swedish registry setting. Operating times, or, in the case of knee surgery, tourniquet times, are associated with the risk of PJI after hip or knee joint replacement,[@Maoz2015; @Willis-Owen2010] and procedure time is one of the risk factors in the Mayo PJI risk score,[@Berbari2012] but we had no information on this potentially important parameter.
Taken together, the inclusion of some or all of the variables mentioned above-unfortunately unavailable to us-might improve the predictive power of PJI risk assessment models. The question of what other variables might be needed to further refine prediction models remains open, but information that is notoriously difficult to obtain within the setting of large register studies, such as smoking status, might be of value. The question of whether there is a genetic predisposition towards developing orthopaedic infections has only been ...

@@

```{r coefs}
tbl_coefs %>% 
  mutate(
    time = replace(as.character(time), duplicated(time), ""),
    beta = round(beta, 2)
  ) %>% 
knitr::kable(
  caption = 'Estimated coefficients (beta) and odds ratios (OR) with 95 % confidence intervals for the reduced models. BMI = body mass index ("under/normal weight as baseline). CNS = central nervous system. Primary osteoarthritis was baseline for diagnosis and female was baseline for sex.'
)
```

>>>>

@@

**RESULTS: ** The crude incidence rate of PJI after THA was `r crude_rate_90d * 100` % within 90, days and `r crude_rate_2y * 100` % within two years in the Swedish derivation cohort. A prediction model for PJI within 90 days combined `r coefs_text["90 days"]`, resulting in an `r get_auc("90 days", "Reduced model", "2")` internally  and **(AUC=XXX, 95% CI: XXX to XXX)** externally. A model for PJI within two years had similar performance internally `r get_auc("2 years", "Reduced model")` and externally **XXX (95% CI: XXX to XXX)**. These models were superior to traditional models based on the American Society for Anesthesiologists (ASA) classification `r get_auc("90 days", "ASA class")`, Charlson `r get_auc("90 days", "Charlson")`, Elixhauser `r get_auc("90 days", "Elixhauser")`, and the Rx Risk V `r get_auc("90 days", "Rx Risk V")` co-morbidity indices.

@@

Periprosthetic joint infection (PJI) is the most devastating of all early complications after total hip arthroplasty (THA), mainly due to its severity in terms of increased mortality, the required number of re-operations, long-term antibiotic treatment, and often persistently impaired quality of life.[@Zimmerli2006; @Kapadia2016; @Shohat2019] Many risk factors for the development of PJI are identified, with anemia, diabetes, and obesity being important, to some extent modifiable risk factors, whereas advanced age and male gender are examples of non-modifiable risk factors.[@Ridgeway2005; @Lai2007; @Malinzak2009; @Ong2009; @Pedersen2010; @Bozic2012; @Jamsen2012; @Rasouli2014a; @Maoz2015]
Individualized PJI risk stratification would be a highly desirable tool in the pre-operative assessment of patients scheduled for THA, but established measures of co-morbidity such as the American Society for Anesthesiologists (ASA) classification or considerably more complex co-morbidity indices cannot be used in this context. Although high ASA class is a risk factor associated with development of PJI, the classification is imprecise, and it was never developed for the purpose of predicting this specific complication.[@Haynes1995; @Berbari2012] 
The Charlson[@Charlson1987] and Elixhauser[@Elixhauser1998] co-morbidity indices are based on `r nrow(coder::charlson)` or `r nrow(coder::elixhauser)` conditions coded by the International Classification of Diseases (ICD), respectively, but they are cumbersome to use in a clinical setting, and they were also not designed for prediction of specific adverse events after THA. The same is true for the Rx Risk V-classification with `r nrow(coder::rxriskv)` conditions codified by the Anatomical Therapeutic Chemical (ATC) Classification.[@Sloan2003; @Pratt2018]
Several attempts have been made at developing individualized risk prediction models for clinical usage, with the Universal American College of Surgeons National Surgical Quality Improvement Program (ACS NSQIP) Surgical Risk Calculator being one of the first.[@Edelstein2015; @Wingert2016a] Several other risk calculators for the risk of PJI after THA have since been derived on US or Australian cohorts, but none has gained general acceptance.[@Wingert2016a; @Bozic2013a; @Inacio2015a] This is mainly due to often limited, sometimes single-center based samples, low  discriminative ability or low accuracy of the prediction models, or the absence of external model validation.
Thus, there is an urgent need for improved prediction models of PJI. We therefore developed a prediction model of PJI within 90 days or two years after THA that is based upon easily accessible data that are available in the setting of clinical decision-making.

@@

The Charlson, Elixhauser and Rx Risk V co-morbidity indices are based on the presence of ICD-10 and ATC codes grouped into main categories (`r nrow(coder::charlson)`, `r nrow(coder::elixhauser)` and `r nrow(coder::rxriskv)` respectively).[@Quan2005; @Pratt2018] Diagnostic categories from all indices were combined to define `r nrow(tab_categorization)` categories of co-morbidity that were chosen in order to be easily identifiable in a clinical setting (Tab. \@ref(tab:tabcategorization)). Some of the combined co-morbidities were observed for less than 10 patients with or without PJI. These diagnoses were not considered as potential predictors to reduce the risk of overfitting to spurious events. PJI was defined by the occurrence of relevant ICD-10 or procedural NOMESCO-code recorded in NPR within 90 days or 2 years (Tab. \@ref(tab:infcodes)), or if the patient was reported to SHAR, within the same time frame, as reoperated due to infection.

@@

For each outcome (PJI within 90 days or 2 years), 1,000 bootstrap samples were drawn from the observed data set.[@Austin2004] Ten-fold cross validation was performed for every bootstrap sample with a range of potential penalty values ($\lambda$:s) in a logistic LASSO model, keeping only $\lambda$:s that minimized the mean cross-validated deviances in each sample and using these $\lambda$:s to estimate model coefficients for each potential predictor. Those coefficients were ranked based on their absolute values as a measure of variable importance. Piece-wise linear regression was applied to find a break-point, and influential predictors with absolute coefficients value above this pointm, scored a point. The whole process was repeated 100 times. Potential predictors that were selected at least once were used in main effects multivariable logistic regression models without penalty and without pre-normalization of numeric variables. Even more parsimonious models with potential predictors chosen at least 33 out of 100 times were evaluated for comparison. Univariable prediction models based on either ASA class, the Charlson, Elixhauser, or the Rx Risk V co-morbidity indices were fitted for comparison. Odds ratios for the final models were estimated with 95 % confidence intervals.

@@

Each developed model was used to predict the probability of PJI within 90 days or two years for patients derived from the SHAR (internal validation). Receiver operating characteristic (ROC) curves and their corresponding areas under those curve (AUC) were used to describe discrimination with 95 % confidence intervals that were based on percentiles from 2,000 non-parametric bootstrap samples. The bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples was used to adjust for optimism, 
but the correction for optimism only affected the third decimals in the presented results, and will therefore not be further discussed. Predicted probabilities and observed proportions were plotted with 95 % confidence intervals to graphically assess model calibration.[@Nattino2016] The derived models were externally validated on a **XXX** cohort, and re-calibration of model intercepts was performed to account for different PJI incidences in Sweden compared to **XXX**. Calibration for this re-calibrated model was illustrated in the same calibration belt plot as after internal calibration.

@@

We built an online web calculator available at https **XXX** to be used in clinical practice. R software (version 3.6.1; R Foundation for Statistical Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr, pROC, decoder, coder and shiny were used.

@@

```{r}
age <- sprintf("%.0f (SD = %.0f)", mean(df$P_Age), sd(df$P_Age))

@@

We included `r nmbr(N_shar)` patients (`r mean(df$P_Sex == "Male") * 100` % males) with a mean age of `r age` years in the derivation cohort from the SHAR (Fig. \@ref(fig:flowchart), left panel). 
`r crude_rate_90d * 100` % (n = `r sum(df$outcome_infection_90d)`) of the patients developed a PJI within 90 days, and `r crude_rate_2y * 100` % (n = `r sum(df$outcome_infection_2y)`) within 2 years (Tab. \@ref(tab:tab1)). The proportion of patients with ASA class III was
`r comorb(df$P_ASA == "III")`; `r comorb(df$CCI_index_quan_original > 0)` had at least one comorbidity according to the Charlson co-morbidity index, `r comorb(df$ECI_index_sum_all > 0)` according to Elixhauser, and `r comorb(df$Rx_index_pratt > 0)` according to Rx Risk V. In addition, **XXX** patients from **XXX** were included as an external validation cohort (Fig. \@ref(fig:flowchart), right panel).

@@

Patients with `r excl_factors` were uncommon, wherefore those co-morbidities were excluded as potentisl predictors prioir to further modeling. ROC-curves (Fig. \@ref(fig:rocs)) and AUC-values (Fig. \@ref(fig:aucci)) for the main and reduced models were no different. For simplicity, we therefore focus on the reduced models. Estimated probability density curves for patients with and without PJI were partially overlapping, although patients with PJI had on average higher predicted probabilites for such event (Fig. \@ref(fig:sep)).

The reduced model for the prediction of PJI within 90 days in the Swedish cohort included `r coefs_text["90 days"]` (Tab. \@ref(tab:coefs)). Predictors associated with a statistically significant risk increase were **secondary osteoarthritis as the underlying reason for THA surgery, male gender, obesity, and the presence of psoriasis, osteoporosis, Paget's or liver disease**. This reduced model for the predicition of PJI within 90 days had an `r get_auc("90 days", "Reduced model", "2")`, whereas ASA class `r get_auc("90 days", "ASA class")`, the Charlson co-morbidity index `r get_auc("90 days", "Charlson")`, the Elixhauser co-morbidity index `r get_auc("90 days", "Elixhauser")`, and Rx Risk V `r get_auc("90 days", "Rx Risk V")` had less discriminative ability (Fig. \@ref(fig:rocs) and \@ref(fig:aucci)). The reduced model had good accuracy as seen by the calibration plot (Fig. \@ref(fig:calibration), left panel).

The reduced model for PJI within two years included the same predictors, with the addition of `r coefs_text["2 years"]` (Tab. \@ref(tab:coefs)). Predictors conferring a statistically significant risk increase were those **mentioned as predictive for PJI within 90 days, excluding osteoporosis or Paget's disease, and additionally included an underlying diagnosis of AVN or inflammatory joint disease, and the presence of  CNS disease**. The reduced model for the prediction of PJI within two years had an `r get_auc("2 years", "Reduced model", "2")`, whereas ASA class `r get_auc("2 years", "ASA class")`, the Charlson co-morbidity index `r get_auc("2 years", "Charlson")`, the Elixhauser co-morbidity index `r get_auc("90 days", "Elixhauser")`, and Rx Risk V `r get_auc("90 days", "Rx Risk V")` again had less discriminative ability (Fig. \@ref(fig:rocs) and \@ref(fig:aucci)). Calibration was good for the reduced model (Fig. \@ref(fig:calibration), right panel).

@@

We found that a multivariable main effects logistic regression model based on some easily identifiable clinical conditions, gender, and BMI was considerably better at predicting PJI within 90 days after THA than models based on either ASA class or the Charlson, Elixhauser or Rx Risk V co-morbidity indices. A similarly developed model to predict PJI within two years included similar predictors as the model developed to predict PJI within 90 days, but added `r coefs_text["2 years"]` to the list of predictor variables. This model also performed better than the established co-morbidity measures in terms of AUC and calibration.

@@

Since the prediction models both for PJI within 90 days and within two years were based on easily accessible information they can be translated into a simple formula $\hat p = 1/(1 + e^{-\hat \alpha - \mathbf{\hat \beta X}})$ where $\hat p$ is the estimated probability, $\hat \alpha$ and $\mathbf{\beta} = \hat\beta_1, \hat\beta_2, \ldots, \hat\beta_k$ are the estimated intercept and coefficients (Tab. \@ref(tab:coefs)) and $\mathbf{X} = X_1, X_2, \ldots, X_k$ are corresponding patient characteristics. For example, `r examples$desc[[1]]` for PJI within 90 days; `r examples$desc[[2]]`.

@@

Patients with primary osteoarthritis seem less prone to developing PJI than patients receiving THA for other reasons, and in agreement with previous observational data we found that both AVN and secondary osteoarthritis conferred a nearly two-fold increased risk of PJI in our prediction models.[@Pedersen2010a; @Dale2012] Rheumatoid arthritis as the diagnosis underlying THA surgery is pointed out as a risk factor for PJI,[@Bozic2012; @Inacio2015a] and this predictor variable was statistically significance in our model of PJI within two years.

Obesity is associated with an increase in the risk of reoperations or adverse events after THA[@Sayed-Noor2019] and with the risk of developing surgical site infections or PJI after total hip or knee arthroplasty.[@Malinzak2009; @Jamsen2012; @Maoz2015; @Everhart2013] Obesity is identified as a risk factor for PJI in a risk calculator derived on two independent US institutional cohorts, which agrees with our identification of obesity as a predictor variable in both models.[@Tan2018] In accordance with observational studies[@Jamsen2012] we found the presence of CNS diseases to be a predictor of PJI within two years, and our rather broad category encompasses both cerebrovascular disease, dementia, hemiplegia, and Parkinson's disease which were all associated with a risk increase in the cited study.

Male gender is associated with an increased risk of developing PJI after arthroplasty surgery,[@Dale2012] and gender is included as a predictor variable in several previously developed risk calculators.[@Wingert2016a; @Tan2018] 
Patients with liver cirrhosis are described to have a more than doubled risk of suffering from PJI within one year after THA, supporting our finding of the presence of liver diseases among our predictor variables.[@Deleuran2015]Taken together, the above cited findings support that the predictor variables in our models seem to be relevant in the context of predicting PJI.

Some previously mentioned risk factors for developing PJI were not selected during our model development. High ASA class is frequently associated with an increased risk of PJI,[@Berbari2012; @Pulido2008; @Namba2012] as are co-morbidity in a broader sense, and, more specifically, the presence of cardiovascular co-morbidity.[@Ong2009; @Higuera2011] In contrast, we found that ASA class alone was an insufficient predictor of PJI, and none of the cardiovascular disease categories remained as a predictor variable in our final main model.

Cancer seems to confer an increased risk of PJI in several observational studies,[@Bozic2012; @Everhart2013; @Poultsides2013] and it is one of the defined co-morbidities in the ACS NSQIP Surgical Risk Calculator,[@Wingert2016a]
20 but this comorbidity did not reach statistical significance in our models.

THA fixation using cement without antibiotics confers an increased risk of PJI,[@Dale2012] but such cement brands were not in clinical use in Sweden during the studied period. We found no indication that the type of fixation was associated with the risk of PJI, and this variable was also not selected as a potential predictor variable in any of the iterations during model development.

Although mentioned as risk factors for surgical site infections after joint arthroplasty,[@Rasouli2014a; @Inacio2015a] prediction models based on the co-morbidity indices developed by Charlson or Elixhauser and the RxRisk V index alone, resulted in poor predictive power. This agrees with previous findings describing that these measures of comorbidity rather poorly predict mortality and patient reported outcomes after THA.[@Bulow2017; @Bulow2019]

@@

Our study is hampered by the limitations inherent to observational research, such as misclassification, and residual confounding due to known confounders that we had no information on, or due to unknown confounders. Importantly, some previously identified risk factors for PJI were not included in our models, mostly because we lacked detailed information on laboratory findings and non-prescribed medications.

The presence of diabetes was not among the predictor variables selected in our models, although it was almost double as frequent among patients with PJI than among those without (data not shown). This is in disagreement with numerous observational studies describing an increased risk of developing PJI after both THA or total knee arthroplasty,[@Pedersen2010; @Maoz2015; @Everhart2013] and diabetes is also found among the risk factors in previously developed risk calculators.[@Wingert2016a; @Tan2018; @Inacio2015] On the other hand, diabetes is not consistently found to be associated with the risk of surgical site infection or PJI after THA surgery.[@Namba2012] Peri-operative hyperglycemia is a predictor of PJI after total joint replacement of the hip or knee, and information on laboratory parameters such as morning blood glucose or HBA1c could have improved the predictive power of our models, but we had no access to such information.[@Mraovic2011; @Iorio2012] In analogy, low preoperative hemoglobin is associated with the risk of developing surgical site infection,[@Rasouli2014a] but although we had access to the ICD codes defining the presence of anemia, we had no information on actual laboratory findings. 
Use of TNF-a blockers may be associated with the risk of developing PJI.[@Gilson2010; @Momohara2011] Similarly, intra-articular steroid application may enhance this risk,[@Cancienne2015] but injections given by physicians would not necessarily be identifiable within a Swedish registry setting. Operating times, or, in the case of knee surgery, tourniquet times, are associated with the risk of PJI after hip or knee joint replacement,[@Maoz2015; @Willis-Owen2010] and procedure time is one of the risk factors in the Mayo PJI risk score,[@Berbari2012] but we had no information on this potentially important parameter. The question of what other variables might be needed to further refine prediction models remains open, but information that is notoriously difficult to obtain within the setting of large register studies, such as smoking status, might be of value.

Taken together, the inclusion of some or all of the variables mentioned above -- unfortunately unavailable to us -- might improve the predictive power of PJI risk assessment models.

@@

```{r coefs}
tbl_coefs_present %>% 
  mutate(
    time = replace(as.character(time), duplicated(time), ""),
    beta = round(beta, 2)
  ) %>% 
knitr::kable(
  caption = 'Estimated coefficients (beta) and odds ratios (OR) with 95 % confidence intervals for the reduced models. BMI = body mass index ("under/normal weight as baseline). CNS = central nervous system. Primary osteoarthritis was baseline for diagnosis and female was baseline for sex.'
)
```
